Sampling requirements to detect ecosystem change

### 2020-10-05 R/manuscript_draft.Rmd
trying out a new col palette

Environmental monitoring is a key component of understanding and managing ecosystems. Given that most monitoring efforts are still expensive and time-consuming, it is essential that monitoring programs are designed to be efficient and effective. In many situations, the expensive part of monitoring is not sample collection, but instead sample processing, which leads to only a subset of the samples being processed. For example, sediment or ice cores can be quickly obtained in the field, but they require weeks or months of processing in a laboratory setting. Standard sub-sampling approaches often involve equally-spaced sampling. We use simulations to show how many samples, and which types of sampling approaches, are the most effective in detecting ecosystem change. We test these ideas with a case study of Cladocera community assemblage indicators reconstructed from a sediment core. We demonstrate that standard approaches to sample processing are less efficient than an iterative approach. For our case study, using an optimal sampling approach would have resulted in savings of 195 person-hours---thousands of dollars in labor costs. We also show that, compared with these standard approaches, fewer samples are typically needed to achieve high statistical power. We explain how our approach can be applied to monitoring programs that rely on video records, eDNA, remote sensing, and other common tools that allow re-sampling.

@@

Environmental monitoring is one of the core components to modern ecosystem research and management [@McDonald-Madden2010a; @White2019e; @Lindenmayer2020a]. Within an adaptive management framework, monitoring is needed for both learning about the system under study and assessing the effectiveness of management interventions [@Lovett2007]. Increasingly, long-term monitoring programs, like the Long Term Ecological Research (LTER) Network in the USA, are becoming available [@Maguran2010a]. However, environmental monitoring is still often expensive and time-consuming, especially when further processing is needed following sample collection [@Zhang2012]. Thus, for many fields there is a disparity between the amount of data that can be acquired and stored, and the ultimate number of samples that can be processed. Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently [@Legg2006a; @McDonald-Madden2010a; @Lengyel2018; @Lindenmayer2020a].

Monitoring programs characteristics must be tightly linked to the questions of interest. For example, @White2019e found that 72\% of vertebrate populations required at least 10 years of monitoring to detect significant changes in the population size over time. The specific number of years required depended on the species biology and the detection method used [@White2019e]. Other work has focused on the frequency of monitoring [@Wauchope2019a], the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a; @Weiser2019a], imperfect detection [@Morant2020], and the costs and benefits of increasing sampling breadth relying on citizen science [@Weiser2020a]. Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection [@Mapstone1995a]. Given limited budgets, monitoring programs need to be designed to be cost-effective [@Caughlan2001a; @Grantham2008b; @Bennett2016a].

@@

Because of new technological advances, there are many data sources that can be derived long after the actual  processes occurred. For example, sediment cores can be retrieved from aquatic ecosystems with little sediment disturbances, such as lakes or lagoon, allowing reconstruction of past ecological communities or conditions [@Cohen2003]. Similarly, environmental samples (e.g. water, soil) can be saved and processed later for composition, including eDNA [@Bohmann2014a]. Likewise, photo- or video-based monitoring can record snapshots of a system and be analyzed later [@OConnell2011a;@Mallet2014a]. In each of these cases, decisions have to made about how much data to extract from the previously collected samples [@Zhang2012]. Should the paleoecological core be analyzed at every centimeter? Should the video be assessed once per minute? As long as processing samples is expensive, these trade-offs will remain.

@@

The random sampling approach involves taking a set number of random points throughout the time series (Fig. \ref{fig:conceptual}c). In the context of sediment cores, this would mean analyzing community composition at random locations along the core. Random sampling is recommended in designs aimed at quantifying the average size of a population (spatial approach) [@Nad2018]. We hypothesize that random sampling will perform the worst in estimating the changepoint. Regular sampling is commonly used (e.g., pigments in @Milan2015) and requires that a set number of samples be taken at regular intervals (Fig. \ref{fig:conceptual}d). Lastly, iterative sampling involves first taking a set number of samples and then iteratively adding samples until a pre-determined level of precision is achieved (Fig. \ref{fig:conceptual}e-g). For each scenario, we begin by sampling the first and last sample to ensure coverage of the whole time period. We describe each approach in more detail in the supplementary material and provide code.

@@

In line with theory on optimal monitoring, we found that the probability of correcting identifying a changepoint decreased with smaller levels of population variability ($\sigma$) and autocorrelation ($\phi$) (Fig. \ref{fig:parameter_sensitivity_plot}). We also found that the probability of correct changepoint detection increases with larger shift sizes, which is essentially the effect size (Fig. \ref{fig:parameter_sensitivity_plot}). There were interaction effects between the variables. For example, autocorrelation was only important if population variability was high (Fig. \ref{fig:parameter_sensitivity_plot}). Thus, the number of samples required to obtain high statistical power (above 0.8) increased with larger population variability, lower autocorrelation, and smaller shift sizes (Fig. \ref{fig:simulation_min_time_plot}). As predicted, iterative sampling performed best, followed by regular and random sampling (Fig. \ref{fig:simulation_min_time_plot}). The distance to the true changepoint, and consequently the minimum number of samples required, was lower for iterative sampling. (Fig. \ref{fig:simulation_min_time_plot})

@@

In a previous study, Cladoceran assemblage was reconstructed continuously with a 1-cm subsampling resolution (2-3 years resolution), from a 74-cm sediment core covering the 1816(±26)—2010 time period [@Bruel2018]. Our objective was to evaluate whether the same changepoints could be identified using less samples. In this previous study, the variability in the community was summarized into independent axis using Detrended Component Analysis [@Hill1980]. Changepoints were then detected on the first component (46% of the total variability) in years 1926, 1946, and 1983. We defined these as the "true" changepoints given they came from an analysis of the complete data. The 1983 changepoint was the largest in magnitude, hence the changepoint we sought to find with our method. We also identified second and third changepoints (Fig. S3).

In line with our simulation approach, we subsampled the full record (74 observations) using the three methods described earlier (random, regular, iterative). These subsamples were from the initial community dataset (Fig. \ref{fig:case_study_iterative_sampling}a). We reduced the dimensionality of the assemblage-level data to an ordination axis using the same method than in the original study, and detected the changepoint on the first component (univariate vector). In the case of the iterative method, a new sample was added and the ordination was run again (Fig. \ref{fig:case_study_iterative_sampling}d). For each of the three methods, we examined the error (difference between the true changepoint and the detected changepoint) when using different numbers of samples.

@@

As another application of our approach, we examined the same sediment core data, but examined total abundance as opposed to community composition (Fig. S4). We tested the three subsampling methods, and it took 11 samples to find the "true" changepoint  (Fig. S5). The initial 5 subsamples analyzed were the same than the subsamples analyzed to answer the question of the change in community  (Fig. \ref{fig:case_study_number_sample}c). The implication is that a very limited number of processed sample can rapidly and efficiently be used to narrow down different questions on a same dataset. <!--However, depending on the metric of interest, the subsampling will target different subsamples to add. In other words, the number of subsamples analyzed can be reduced, especially in cases where the question is well tailored from the beginning.-->

@@

Our approach goes beyond just paleoecological analyses. Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019e; @White2020]. The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost. Our specific approach applies to situations where more subsamples can be added, or processed, after the dynamics occurred [@Zhang2012]. It corresponds very well to paleoecological data: samples are taken long after the phenomenon of interest occurred, and allows subsampling at finer or rougher intervals [@Wingard2017]. However, both different types of data and different questions than those used here can be addressed with the same approach. Suppose instead that the goal was to detect a change in relative abundance over time with video-based approaches. It is often not practical to watch entire videos, so it can be useful to choose strategic time-points that would address a specific question of interest. Using an interval sampling approach, one could take a fixed number of samples to start. The trend over time from simple linear regression could be taken. Then, samples can be taken at random locations one-by-one and to see which samples have the largest effect on the trend estimates. If a particular sample has a large effect on the trend, then it would be best to choose another nearby sample. Sampling would continue until the trend estimate reached an equilibrium. Thus, the iterative sampling approach is particularly relevant to data sources where additional samples can be taken long after the initial dynamics. These approaches would also be appropriate for environmental samples, such as water or soil, that can be analyzed later or eDNA that can be extracted from previously-collected samples [@Bohmann2014a].

Our approach is applicable to a wide range of systems and questions, but it does have limitations. When less resources are needed for sample analysis, as opposed to collection, investigators will likely be able to process every sample, and analyzing all samples to obtain a whole picture may be preferred. We note that if resources need to be saved by collecting less samples in the first place, then regular sampling performs better than random sampling (Figs. \ref{fig:simulation_approaches_plot}, S1, S2). Another example where our method is less useful is when addressing questions that require a continuous time series, or at least a regular sampling interval. For example, continuous, high-resolution subsampling of a time-series is generally required to detect critical slowing-down or early warning of shifts [@Frossard2015; @Doncaster2016].

@@

mypal <- wes_palette("GrandBudapest2", max(vr_out_order$count),type = "continuous")

>>>>

Environmental monitoring is a key component of understanding and managing ecosystems. Given that most monitoring efforts are still expensive and time-consuming, it is essential that monitoring programs are designed to be efficient and effective. In many situations, the expensive part of monitoring is not sample collection, but instead sample processing, which leads to only a subset of the samples being processed. For example, sediment or ice cores can be quickly obtained in the field, but they require weeks or months of processing in a laboratory setting. Standard sub-sampling approaches often involve equally-spaced sampling \textcolor{red}{on depth}.\linelabel{text:depth} We use simulations to show how many samples, and which types of sampling approaches, are the most effective in detecting ecosystem change. We test these ideas with a case study of Cladocera community assemblage indicators reconstructed from a sediment core. We demonstrate that standard approaches to sample processing are less efficient than an iterative approach. For our case study, using an optimal sampling approach would have resulted in savings of 195 person-hours---thousands of dollars in labor costs. We also show that, compared with these standard approaches, fewer samples are typically needed to achieve high statistical power. We explain how our approach can be applied to monitoring programs that rely on video records, eDNA, remote sensing, and other common tools that allow re-sampling.

@@

Environmental monitoring is one of the core components to modern ecosystem research and management [@McDonald-Madden2010a; @White2019e; @Lindenmayer2020a]. Within an adaptive management framework, monitoring is needed for both learning about the system under study and assessing the effectiveness of management interventions [@Lovett2007]. Increasingly, long-term monitoring programs, like the Long Term Ecological Research (LTER) Network in the USA, are becoming available [@Maguran2010a]. \linelabel{text:remote1} However, environmental monitoring \textcolor{red}{can still be logistically difficult, expensive, and time-consuming, especially} when further processing is needed following sample collection [@Zhang2012]. Thus, for many fields there is a disparity between the amount of data that can be acquired and stored, and the ultimate number of samples that can be processed. Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently [@Legg2006a; @McDonald-Madden2010a; @Lengyel2018; @Lindenmayer2020a].

\linelabel{text:program} \textcolor{red}{Monitoring program characteristics must be tightly linked to the questions of interest.} For example, @White2019e found that 72\% of vertebrate populations required at least 10 years of monitoring to detect significant changes in the population size over time. The specific number of years required depended on the species biology and the detection method used [@White2019e]. \linelabel{text:newcitation} \textcolor{red}{In addition, the sampling effort required differs depending on the question. For example, questions regarding phenology would require many sampling points within a season and across years} [@Filippa2015a, @White2020e]. Other work has focused on the frequency of monitoring [@Wauchope2019a], the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a; @Weiser2019a], imperfect detection [@Morant2020], and the costs and benefits of increasing sampling breadth relying on citizen science [@Weiser2020a]. Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection [@Mapstone1995a]. Given limited budgets, monitoring programs need to be designed to be cost-effective [@Caughlan2001a; @Grantham2008b; @Bennett2016a].

@@

Because of new technological advances, there are many data sources that can be derived long after the actual  processes occurred. For example, sediment cores can be retrieved from aquatic ecosystems with little sediment disturbances, such as lakes or lagoon, allowing reconstruction of past ecological communities or conditions [@Cohen2003]. Similarly, environmental samples (e.g. water, soil) can be saved and processed later for composition, including eDNA [@Bohmann2014a]. Likewise, photo- or video-based monitoring can record snapshots of a system and be analyzed later [@OConnell2011a;@Mallet2014a]. In each of these cases, decisions have to made about how much data to extract from the previously collected samples [@Zhang2012]. Should the paleoecological core be analyzed at every centimeter? Should the video be assessed once per minute if automated tools are not available? As long as processing samples is expensive, these trade-offs will remain.

@@

The random sampling approach involves taking a set number of random points throughout the time series (Fig. \ref{fig:conceptual}c). In the context of sediment cores, this would mean analyzing community composition at random locations along the core. Random sampling is recommended in designs aimed at quantifying the average size of a population (spatial approach) [@Nad2018]. We hypothesize that random sampling will perform the worst in estimating the changepoint. Regular sampling is commonly used (e.g., pigments in @Milan2015) and requires that a set number of samples be taken at regular intervals (Fig. \ref{fig:conceptual}d). Lastly, iterative sampling involves first taking a set number of samples (i.e., regular sampling) and then iteratively adding samples until a pre-determined level of precision is achieved (Fig. \ref{fig:conceptual}e-g). \linelabel{text:iterative2} \textcolor{red}{In the context of a changepoint, this means adding a new sample near the best estimate for the changepoint and iteratively updating the estimate (Fig. \ref{fig:conceptual}e-g).} For each scenario, we begin by sampling the first and last sample to ensure coverage of the whole time period. \textcolor{red}{We describe each approach in more detail in the supplementary material and provide code.} \linelabel{text:iterative1}

@@

In line with theory on optimal monitoring, we found that the probability of correcting identifying a changepoint decreased with smaller levels of population variability ($\sigma$) and autocorrelation ($\phi$) (Fig. \ref{fig:parameter_sensitivity_plot}). We also found that the probability of correct changepoint detection increases with larger shift sizes, which is essentially the effect size (Fig. \ref{fig:parameter_sensitivity_plot}). \linelabel{text:detection1} \textcolor{red}{In line with theory, if the shift size was small, and thus there was no changepoint, there was almost never enough power to detect a shift. Similar results have been shown for the detection of linear trends} [@White2019e].  There were interaction effects between the variables. For example, autocorrelation was only important if population variability was high (Fig. \ref{fig:parameter_sensitivity_plot}). Thus, the number of samples required to obtain high statistical power (above 0.8) increased with larger population variability, lower autocorrelation, and smaller shift sizes (Fig. \ref{fig:simulation_min_time_plot}). As predicted, iterative sampling performed best, followed by regular and random sampling (Fig. \ref{fig:simulation_min_time_plot}). The distance to the true changepoint, and consequently the minimum number of samples required, was lower for iterative sampling. (Fig. \ref{fig:simulation_min_time_plot})

@@

In a previous study, Cladoceran assemblage was reconstructed continuously with a 1-cm subsampling resolution (2-3 years resolution), from a 74-cm sediment core covering the 1816(±26)—2010 time period [@Bruel2018]. Our objective was to evaluate whether the same changepoints could be identified using fewer samples.\linelabel{text:fewer} In this previous study, the variability in the community was summarized into independent axis using Detrended Component Analysis [@Hill1980]. Changepoints were then detected on the first component (46% of the total variability) in years 1926, 1946, and 1983. We defined these as the "true" changepoints given they came from an analysis of the complete data. The 1983 changepoint was the largest in magnitude, hence the changepoint we sought to find with our method. We also identified second and third changepoints (Fig. S3).

In line with our simulation approach, we subsampled the full record (74 observations) using the three methods described earlier (random, regular, iterative). These subsamples were from the initial community dataset (Fig. \ref{fig:case_study_iterative_sampling}a). We reduced the dimensionality of the assemblage-level data to an ordination axis using the same method in the original study, and detected the changepoint on the first component (univariate vector). \linelabel{text:iterative} \textcolor{red}{In the case of the iterative method, a new sample was added in the temporal region closest to the estimate for the changepoint until a more precise estimate was obtained} (Fig. \ref{fig:case_study_iterative_sampling}d). For each of the three methods, we examined the error (difference between the true changepoint and the detected changepoint) when using different numbers of samples.

@@

\linelabel{text:detection2} \textcolor{red}{To show the generality of our approach}, we examined the same sediment core data, but examined total abundance as opposed to community composition (Fig. S4). We tested the three subsampling methods, and it took 11 samples to find the "true" changepoint  (Fig. S5). The initial 5 subsamples analyzed were the same than the subsamples analyzed to answer the question of the change in community  (Fig. \ref{fig:case_study_number_sample}c). The implication is that a very limited number of processed sample can rapidly and efficiently be used to narrow down different questions on a same dataset. \textcolor{red}{In addition, we examined the same abundance time series, but with linear and generalized additive models (GAMs)}. In line with findings from} @White2019e, \textcolor{red}{we found that the flexibility of GAMs allow the use of less samples (Figs. S6, S7).}

<!--However, depending on the metric of interest, the subsampling will target different subsamples to add. In other words, the number of subsamples analyzed can be reduced, especially in cases where the question is well tailored from the beginning.-->

@@

Our approach goes beyond just paleoecological analyses. Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019e; @White2020] or \textcolor{red}{logistically difficult, such as with remote field sites}.\linelabel{text:remote} The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost. \linelabel{text:newcitation1} \textcolor{red}{Similarly, in order to evaluate ecosystem phenology,} @Filippa2015a \textcolor{red}{used similar techniques to show the effect of not only different levels of sampling, but also the effect of using different indices altogether.} Our specific approach applies to situations where more subsamples can be added, or processed, after the dynamics occurred [@Zhang2012]. It corresponds very well to paleoecological data: samples are taken long after the phenomenon of interest occurred, and allows subsampling at finer or rougher intervals [@Wingard2017]. However, both different types of data and different questions than those used here can be addressed with the same approach. Suppose instead that the goal was to detect a change in relative abundance over time with video-based \textcolor{red}{approaches where automated techniques are not possible}. It is often not practical to watch entire videos, so it can be useful to choose strategic time-points that would address a specific question of interest. Using an interval sampling approach, one could take a fixed number of samples to start. The trend over time from simple linear regression could be taken. Then, samples can be taken at random locations one-by-one and to see which samples have the largest effect on the trend estimates. If a particular sample has a large effect on the trend, then it would be best to choose another nearby sample. Sampling would continue until the trend estimate reached an equilibrium. Thus, the iterative sampling approach is particularly relevant to data sources where additional samples can be taken long after the initial dynamics. \linelabel{text:otherapproaches} \textcolor{red}{These approaches would also be appropriate for environmental samples, such as water or soil, that can be analyzed later or eDNA that can be extracted from previously-collected samples} [@Bohmann2014a]. \textcolor{red}{The same approaches could also be used in studying evidence of stressers found in tree rings or coral reef cores} [@Carilli2012a; @Pretzsch2013a], \textcolor{red}{pollen in aerial traps or sediment cores to address phenological questions} [@Haselhorst2013a], \textcolor{red}{and many others situations described in} @White2020. \linelabel{text:otherapproaches1}

Our approach is applicable to a wide range of systems and questions, but it does have limitations. When less resources are needed for sample analysis, as opposed to collection, investigators will likely be able to process every sample, and analyzing all samples to obtain a whole picture may be preferred. We note that if resources need to be saved by collecting less samples in the first place, then regular sampling performs better than random sampling (Figs. \ref{fig:simulation_approaches_plot}, S1, S2). \linelabel{red}{text:limits} \textcolor{red}{Our specific iterative approach for detecting changepoints is also not appropriate for systems where changepoints are not expected. Instead, more flexible models (e.g., GAMs) might be more appropriate (Fig. S7).} Another example where our method is less useful is when addressing questions that require a continuous time series, or at least a regular sampling interval. For example, continuous, high-resolution subsampling of a time-series is generally required to detect critical slowing-down or early warning of shifts [@Frossard2015; @Doncaster2016].

@@

mypal <- wes_palette("GrandBudapest2", max(vr_out_order$count),type = "continuous")
#mypal <-  c("#4C4F54", "#D57A5D", "#AE5654", "#D18437", "#9EC6DD")
